---
title: "TdS Toblach Freestyle Sprint Predictions"
author: "Joran Elias"
date: "2023-12-30"
categories: [tour de ski,sprint,predictions,freestyle,toblach]
draft: false
---

```{r pkgs}
library(tidyverse)
library(statskier2)
library(gt)
library(gtExtras)
library(ggpubr)
library(arrow)

theme_set(theme_bw())
```

I wasn't super organized last night, so you'll have to just trust me that I 
generated these predictions without feeding in the actual results. Honest!

I did tweak it a little bit to weight recent results more heavily, but I don't
think it made an enormous difference (yet).

```{r pred-data}
location <- "Toblach"
tech <- "Freestyle"
mens_preds <- 
  readr::read_csv(file = '2023-12-30_men_fr_spr_toblach.csv',
                  show_col_types = FALSE)
womens_preds <- 
  readr::read_csv(file = '2023-12-30_wom_fr_spr_toblach.csv',
                  show_col_types = FALSE)
```

### Men's Predictions
```{r tbl-men}
mens_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = glue::glue("{loc} {tech} Sprint",loc = location,tech = tech),
             subtitle = "Men") |> 
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank)
```

# Women's Predictions
```{r tbl-wom}
womens_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = glue::glue("{loc} {tech} Sprint",loc = location,tech = tech),
             subtitle = "Women") |>
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank)
```

### Update
With a few sprints in the books, let's take a look.
```{r db-con}
statskier_connect()
v_spr_maj_int <- dplyr::tbl(..statskier_pg_con..,'v_sprint_maj_int')
```

```{r race-data}
spr_res <- v_spr_maj_int |> 
  filter(eventid %in% c(19376,19377)) |> 
  collect()

#fis_pts <- readRDS(file = '~/Documents/code/official-fis-points-lists/output/fis_pts_list.rds') |> 
fis_pts <- arrow::read_parquet(file = '~/Documents/code/official-fis-points-lists/output/fis_pts_list_2023-12-28.parquet') |> 
  select(listid = Listid,
         #listname = Listname,
         listnum = ListNum,
         compid = Competitorid,
         year = year,
         calc_date = Calculationdate,
         spr_points = SPpoints) |> 
  filter(year >= 2020) |> 
  distinct()

men_comb <- spr_res |> 
  filter(gender == 'Men') |> 
  select(date,compid,rank) |> 
  mutate(date = as.Date(date)) |> 
  full_join(mens_preds,by = 'compid') |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(
    fis_pts_rank = rank(spr_points,ties.method = 'min')) |> 
  arrange(rank)

wom_comb <- spr_res |> 
  filter(gender == 'Women') |> 
  select(date,compid,rank) |> 
  mutate(date = as.Date(date)) |>
  full_join(womens_preds,by = 'compid') |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(
    fis_pts_rank = rank(spr_points,ties.method = 'min')) |> 
  arrange(rank)
```

Not much difference between the two in this race.

```{r men-perf}
men_comb |> 
  select(compid,rank,page_rank_pos,fis_pts_rank) |> 
  pivot_longer(cols = c(page_rank_pos,fis_pts_rank),names_to = 'grp',values_to = 'val') |> 
  mutate(grp = if_else(grp == 'fis_pts_rank','FIS Dst Pts List','PR Method')) |> 
  ggplot(data = _,aes(x = rank,y = val)) + 
  facet_wrap(~grp,ncol = 1) + 
  geom_point() + 
  geom_abline(slope = 1,intercept = 0,color = 'blue') + 
  stat_cor(aes(label = after_stat(..r.label..))) +
  ggtitle(label = glue::glue("Men's {loc} {tech} Sprint Predicted vs True Athlete Finishing Order",loc = location,tech = tech),
          subtitle = "Athletes ordered by most recent FIS Spr pts list (top) & my PR method (bottom)") +
  labs(x = 'True Finishing Place',y = 'Predicted Finishing Place')
```

```{r wom-perf}
wom_comb |> 
  select(compid,rank,page_rank_pos,fis_pts_rank) |> 
  pivot_longer(cols = c(page_rank_pos,fis_pts_rank),names_to = 'grp',values_to = 'val') |> 
  mutate(grp = if_else(grp == 'fis_pts_rank','FIS Dst Pts List','PR Method')) |> 
  ggplot(data = _,aes(x = rank,y = val)) + 
  facet_wrap(~grp,ncol = 1) + 
  geom_point() + 
  geom_abline(slope = 1,intercept = 0,color = 'blue') + 
  stat_cor(aes(label = after_stat(..r.label..))) +
  ggtitle(label = glue::glue("Women's {loc} {tech} Sprint Predicted vs True Athlete Finishing Order",loc = location,tech = tech),
          subtitle = "Athletes ordered by most recent FIS Spr pts list (top) & my PR method (bottom)") +
  labs(x = 'True Finishing Place',y = 'Predicted Finishing Place')
```

