---
title: "All Time Best Distance Skiers"
author: "Joran Elias"
date: "2024-03-11"
categories: [world cup,olympics,world championships,distance]
draft: false
execute:
  echo: false
  message: false
  warning: false
---

```{r pkgs}
library(tidyverse)
library(statskier2)
library(geomtextpath)
library(ggthemes)
library(ggrepel)
library(ggtext)
library(gt)

theme_set(theme_minimal())

statskier_connect()
```

```{r data}
v_dst_maj_int <- dplyr::tbl(..statskier_pg_con..,"v_distance_maj_int")
dst_maj_int <- v_dst_maj_int |> 
  filter(!is.na(time) & !is.na(rank) & season >= "1979-1980") |> 
  collect() |> 
  arrange(eventid,time)

dst_maj_int <- 
  dst_maj_int |> 
  mutate(
    format_red = if_else(format %in% c("Mass","Skiathlon"),"Mass/Skiathlon",format)
  ) |> 
  mutate(
    .by = c(gender,format_red),
    pb_ecdf = cume_dist(pb),
    pbm_ecdf = cume_dist(pbm)
  )

ecdf_funs <- 
  dst_maj_int |> 
  group_by(gender,format_red) |> 
  summarise(
    pb_ecdf = list(ecdf(pb)),
    pbm_ecdf = list(ecdf(pbm))
  )
```

Putting skier's performances in historical context is something I've written about a lot,
in a variety of forms. Most recently I looked at simple [rates of winning & podiums](https://statisticalskier.netlify.app/posts/2024-02-14-podium-counts-pcts/podium_counts_pcts) as
a way of quantifying some of the best skiers in history.

Another way to slice that, focusing only on distance racing, is to focus more on
percent back type measures. The two I like to use are the traditional percent back
from the race winner (PB) and the percent "back" from the median race finisher (PBM).

The former is what FIS points are derived from, the latter is an alternative that
I like to use from time to time. They are measuring slightly different things,
so they each have their own strengths & weaknesses.

Let's first look at the distribution of PB values for the last 45 years or so.

```{r pb-distributions}
#| fig-width: 8
#| fig-height: 5
dst_maj_int |> 
  filter(format != "Pursuit") |> 
  ggplot(
    data = _,
    aes(x = pb,color = format_red,group = format_red)
  ) + 
  facet_wrap(~gender,ncol = 1) +
  geom_textdensity(aes(label = format_red),n = 2048,hjust = 0.075,size = 10 / .pt,linewidth = 0.8) + 
  scale_x_continuous(labels = scales::label_percent()) + 
  scale_color_tableau(guide = "none") +
  labs(x = "Percent from race winner",y = NULL) +
  ggtitle(
    label = "Distributions of the percent back from the race winner",
    subtitle = "All WC, OWG & WSC races since 1980 (pursuit starts excluded)"
  ) +
  coord_cartesian(xlim = c(0,0.25)) + 
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 42 / .pt,face = "bold"),
    axis.text.y = element_blank(),
    strip.text = element_text(size = 30 / .pt)
  )
```

As usual, I'm excluding pursuit starts entirely, as I don't really consider them 
standalone races that can be meaningfully compared to any others. I am including 
"pursuit break" races, which is an old format primarily from the 90's which was
the combined time of a classic & skate race separated by a day. They were sort 
of like modern skiathlons but with a day in between the classic & skate portions.

The standard stuff jumps out here:

  - Men's mass starts are much more likely to generate very low PBs, as the races
  tend to be very slow with a sprint finish
  - Women's mass starts also are more likely to have low PBs, but to much lesser
  extent
  - In general, most men's PBs are between 0-10%, with the most common around 5%
  (mass starts excepted)
  - In general, most women's PBs are between 0-15%, with the most common PBs 
  tending to fall closer to 7.5%
  
Keep in mind that these PB distributions are across *all* skiers, so it includes
the folks finishing last. I'm only plotting the left portion of the distribution,
in reality it extends way out to the right.

Next, let's look at PBM:

```{r pbm-distributions}
#| fig-width: 8
#| fig-height: 5
dst_maj_int |> 
  filter(format != "Pursuit") |> 
  ggplot(
    data = _,
    aes(x = pbm,color = format_red,group = format_red)
  ) + 
  facet_wrap(~gender,ncol = 1) +
  geom_textdensity(aes(label = format_red),n = 2048,hjust = 0.125,size = 10 / .pt,linewidth = 0.8) + 
  scale_x_continuous(labels = scales::label_percent()) + 
  scale_color_tableau(guide = "none") +
  labs(x = "Percent from median race finisher",y = NULL) +
  ggtitle(
    label = "Distributions of the percent back from median race finisher",
    subtitle = "All WC, OWG & WSC races since 1980 (pursuit starts excluded)"
  ) +
  coord_cartesian(xlim = c(-0.1,0.15)) + 
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 42 / .pt,face = "bold"),
    axis.text.y = element_blank(),
    strip.text = element_text(size = 30 / .pt)
  )
```

Since our reference point is the median race finisher, these values don't have a 
hard stop at 0%. The top half of a race will have negative PBMs and the bottom half
will have positive ones. In theory, the race winner could have an arbitrarily low
PBM.

That's one of the advantages of PBM, as it allows for an expression of the margin 
of victory of the race winner. For skiers like Therese Johaug who when she won 
would often win by very large margins, PBM will reflect that, whereas it would get
lost in the usual PB measure.

On the other hand, with PB, we know that 0% always means you won the race. With
PBM you could have a value of -1% or -5% and they might both be races you won. That
information just isn't stored in PBM at all.

The other well known problem with PB (much less so with PBM) is that as we mentioned
above it can mean very different things in different race formats. A PB of 1% has 
a very different meaning for an interval start race versus a mass start in terms
of how many people finished ahead of or behind you.

One simple way to address this is to convert all the PB & PBM values to percentiles.
So instead of a PB of 1%, we'd say that for a men's interval start race it equates 
to a percentile of `r round(1 - ecdf_funs$pb_ecdf[[1]](0.01),2)`, i.e. that
result is in the 96th percentile for men's interval start races. Think of it like
the results you might get on a standardized test saying that your test results
are in the 96th percentile.

```{r pb-pbm-translation-vals}
tbl_vals <- 
  ecdf_funs |> 
  mutate(
    pb01 = map(pb_ecdf,.f = ~do.call(.x,list(0.01))),
    pb05 = map(pb_ecdf,.f = ~do.call(.x,list(0.05))),
    pb10 = map(pb_ecdf,.f = ~do.call(.x,list(0.1))),
    pbm01 = map(pbm_ecdf,.f = ~do.call(.x,list(-0.05))),
    pbm05 = map(pbm_ecdf,.f = ~do.call(.x,list(-0.025))),
    pbm10 = map(pbm_ecdf,.f = ~do.call(.x,list(-0.01)))
  ) |> 
  select(-ends_with("ecdf")) |> 
  filter(format_red != "Pursuit") |> 
  tidyr::unnest(cols = starts_with("pb")) |> 
  mutate(
    across(.cols = starts_with("pb"),.fns = ~1 - .x)
  )
```

Here's a simple table summarising some example translations between PB & PBM
values (column headers) and their corresponding percentiles (table values):

```{r pb-pbm-translation-tbl}
tbl_vals |> 
  group_by(gender) |> 
  gt(row_group_as_column = TRUE) |>
  tab_options(table.width = pct(100)) |> 
  cols_label(
    format_red = "Race format",
    pb01 = "1%",
    pb05 = "5%",
    pb10 = "10%",
    pbm01 = "-5%",
    pbm05 = "-2.5%",
    pbm10 = "-1%"
  ) |> 
  tab_spanner(
    label = "PB",
    columns = matches("^pb[0-9]{2}")
  ) |> 
  tab_spanner(
    label = "PBM",
    columns = matches("^pbm[0-9]{2}")
  ) |> 
  fmt_percent(columns = everything(),decimals = 1) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_column_labels(),cells_column_spanners(),cells_row_groups())
  ) |> 
  tab_style(
    style = cell_borders(sides = "right"),
    locations = cells_body(columns = 5)
  )
  
```

If you do that for all the PB & PBM values you now have everything on a common
scale that means closer to the same thing across all race formats.

If you go even further and summarise each skier by their median PB & PBM 
percentiles you can plot the distribution of all skiers on these two measures:

```{r skier-ecdf-stats}
skier_ecdf <- 
  dst_maj_int |> 
  filter(format != "Pursuit") |> 
  summarise(
    pb_q25 = quantile(pb_ecdf,0.25),
    pb_q50 = quantile(pb_ecdf,0.50),
    pb_q75 = quantile(pb_ecdf,0.75),
    pbm_q25 = quantile(pbm_ecdf,0.25),
    pbm_q50 = quantile(pbm_ecdf,0.50),
    pbm_q75 = quantile(pbm_ecdf,0.75),
    comb_rank = sqrt(pb_q50^2 + pbm_q50^2),
    n = n(),
    last_race_date = max(date),
    .by = c(gender,compid,name)
  ) |> 
  mutate(
    active = if_else(last_race_date >= "2023-10-01" | grepl("BOLSHUNOV",name),"Active","Not Active")
  ) |> 
  arrange(comb_rank)
```

```{r skier-joint-distribution}
#| fig-width: 8
#| fig-height: 5
skier_ecdf |> 
  filter(n >= 30) |> 
  ggplot(
    data = _,
    aes(x = 1 - pb_q50,y = 1 - pbm_q50)
  ) + 
  facet_wrap(~gender,ncol = 2) +
  geom_density_2d_filled(show.legend = FALSE,bins = 15) + 
  annotate(
    geom = "text",
    x = 0.9,
    y = 0.9,
    label = "Best skiers",
    color = "white",
    vjust = 0,
    hjust = 0.75,
    size = 9 / .pt
  ) +
  annotate(
    geom = "text",
    x = 0.1,
    y = 0.1,
    label = "'Worst' skiers",
    color = "white",
    vjust = 1,
    hjust = 0.25,
    size = 9 / .pt
  ) +
  scale_fill_viridis_d(option = "plasma") + 
  scale_x_continuous(labels = scales::label_percent()) + 
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    x = "Median percent back from the race winner",
    y = "Median percent back from the median finisher"
  ) + 
  ggtitle(
    label = "Joint distribution of skiers' % back from the winner & median finisher",
    subtitle = "Only skiers with a minimum of 30 races"
  ) +
  theme(
    panel.grid = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(size = 42 / .pt,face = "bold"),
    strip.text = element_text(size = 30 / .pt)
  )
```

The best skiers are in the upper right, with very high percentiles on both measures.
The "worst" skiers (but still likely faster than all of us) are in the lower left 
with very low percentiles on both measures.

The two measures are highly correlated, as we'd expect, but the degree that these
distributions extend outward from a perfect diagonal line indicates there is 
some variability between to two.

Now let's focus in on the skiers that comprise the upper right of both distributions
here by plotting the raw data for skiers that would fall in the upper right:

```{r best-skiers}
#| fig-width: 8
#| fig-height: 6
skier_ecdf |> 
  filter(n >= 30) |> 
  slice_min(order_by = comb_rank,n = 30,by = c(gender)) |> 
  ggplot(
    data = _,
    aes(x = 1 - pb_q50,y = 1 - pbm_q50,label = name)
  ) + 
  facet_wrap(~gender,ncol = 2) +
  geom_abline(intercept = 0,slope = 1,color = "blue",alpha = 0.5) +
  geom_point(aes(color = active),size = 0.75) +
  geom_text_repel(aes(color = active),size = 6 / .pt) + 
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_percent()) + 
  scale_color_manual(values = c("Active" = "salmon","Not Active" = "black"),guide = "none") +
  labs(
    x = "Median PB",
    y = "Median PBM",
    color = NULL
  ) + 
  ggtitle(
    label = "Top distance skiers from 1980-2024",
    subtitle = "Minimum of 30 races, currently active skiers in <span style = 'color:#FA8072;'>red</span>"
  ) + 
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 42 / .pt,face = "bold"),
    strip.text = element_text(size = 28 / .pt),
    plot.subtitle = element_markdown()
  )
```

I've included Alexander Bolshunov in the "active skier" group, since presumably
he will continue racing if certain world events magically resolve 
themselves. Also, I will let you supply your own commentary on people shown 
here that may or may not have been chemically enhanced.

To help translate this plot, let's focus just on Gunde Svan. What this is saying
is that Svan's *median* PB & PBM across all his races was roughly in the 97th
percentile across all skiers. His "typical" race was a better result than 97% of
all other skier's results *ever*.

A few things I'd note about this collection of the "best" distance skiers:

  - There are considerably more active men in this group than active women
  - Marit Björgen does not fare nearly as well here as I'd expected
  - There's a distinct group of men (Svan, Dæhlie, Wassberg, Bolshunov & Smirnov)
  that are clearly separated from the rest
  - Some skiers do better on one measure than the other; for example, Petter Northug
  does better on PB than PBM, which is probably a reflection of the fact that he
  was particularly known for winning mass start sprint finishes, so he has lots of
  wins in that measure, but the victory margins are small