---
title: "Goms Freestyle Sprint Predictions"
author: "Joran Elias"
date: "2024-01-26"
categories: [world cup,sprint,predictions,classic,oberhof]
draft: false
---

```{r pkgs}
library(tidyverse)
library(statskier2)
library(gt)
library(gtExtras)
library(ggpubr)
library(arrow)

theme_set(theme_bw())
```

Freestyle sprints in Goms should go well for Jessie, you'd think, but my little
algorithm is sticking with a Swedish sweep.

Picking the Valnes to win for the men seems a pretty easy call, but the algorithm
definitely is still weighting pre-illness Klæbo pretty heavily. He had a rough 
weekend last time, but he's got to bounce back eventually you'd think.

```{r pred-data}
location <- "Goms"
tech <- "Freestyle"
mens_preds <- 
  readr::read_csv(file = '2024-01-27_men_fr_spr_goms.csv',
                  show_col_types = FALSE)
womens_preds <- 
  readr::read_csv(file = '2024-01-27_wom_fr_spr_goms.csv',
                  show_col_types = FALSE)
```

### Men's Predictions
```{r tbl-men}
mens_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = glue::glue("{loc} {tech} Sprint",loc = location,tech = tech),
             subtitle = "Men") |> 
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank)
```

# Women's Predictions
```{r tbl-wom}
womens_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = glue::glue("{loc} {tech} Sprint",loc = location,tech = tech),
             subtitle = "Women") |>
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank)
```

