---
title: "Cumulative Win Percentages = GoAT?"
author: "Joran Elias"
date: "2023-11-30"
categories: [world cup]
draft: false
---

```{r pkgs,message=FALSE}
library(tidyverse)
library(statskier2)
library(readxl)

cap <- 'statisticalskier.netlify.app'

theme_set(theme_bw())
statskier_connect()
```

I was chatting with some friends of moderately advanced years today about how
people anointing Johannes Klaebo as the GoAT should get off our lawn. Or at 
least get off Bjorn Daehlie or Marit Bjoergen's lawn.

It was suggested that since the number of available World Cup starts has shifted
pretty dramatically over the years, a decent way to compare skiers would be
cumulative win percentage by age. I'm going to throw in cumulative podium percentage
in there as well just for additional perspective.

I'm taking the win/podium percentage across both sprint & distance events, but 
I'm excluding the modern pursuit races entirely, as "winning" them isn't really
comparable to anything we're terribly interested in.

The other important note is that we should consider this as a (nearly) complete
look from the late 80's forward. So I'm capturing the skiers of the Daehlie, 
Vaelbe generation forward, but my data isn't complete enough from the rest of the
80's back to examine them.

```{r ath-data}
v_dst_maj_int <- dplyr::tbl(..statskier_pg_con..,'v_distance_maj_int')
v_spr_maj_int <- dplyr::tbl(..statskier_pg_con..,'v_sprint_maj_int')

dst <- v_dst_maj_int |> 
  filter(!is.na(rank) & format != 'Pursuit') |> 
  collect() |> 
  mutate(dst_spr = 'Distance')
spr <- v_spr_maj_int |> 
  filter(!is.na(rank)) |> 
  collect() |> 
  mutate(dst_spr = 'Sprint')
```

```{r early}
early_vaelbe <- readxl::read_xlsx(
  path = 'early_vaelbe.xlsx',
  sheet = 'vaelbe') |> 
  mutate(gender = 'Women',
         compid = 63291,
         eventid = -(1:n()))
early_daehlie <- readxl::read_xlsx(
  path = 'early_vaelbe.xlsx',
  sheet = 'daehlie') |> 
  mutate(gender = 'Men',
         compid = 12150,
         eventid = -(1:n()))
```

### Cumulative Win Percentage
Here's cumulative win percentage among World Cup, WSC, OWG and TdS races,
sprint & distance. Maybe the Klaebo stans should, in fact, remain on my lawn.

But Elena Vaelbe probably wins here. (Y'all can argue about doping elsewhere.)

```{r dst-plot}
dst_spr <- 
  bind_rows(
    dst |> select(season,date,eventid,dst_spr,fisid,compid,name,gender,age,rank),
    spr |> select(season,date,eventid,dst_spr,fisid,compid,name,gender,age,rank)
  ) |> 
  bind_rows(early_daehlie) |> 
  bind_rows(early_vaelbe)

started100 <- dst_spr |> 
  summarise(
    n = n_distinct(eventid),
    .by = 'compid'
  ) |> 
  filter(n >= 100)

dst_spr_100 <- dst_spr |> 
  semi_join(started100,by = 'compid')
```

```{r win-pct}
win_pct <- dst_spr_100 |> 
  group_by(gender,season,compid) |> 
  group_by(compid) |> 
  arrange(date) |> 
  mutate(age = timetk::ts_impute_vec(x = age,period = 1)) |> 
  tidyr::fill(name,.direction = 'downup') |> 
  mutate(cum_win_pct = cummean(rank == 1),
         cum_pod_pct = cummean(rank <= 3)) |> 
  ungroup()
```

```{r,cum-win-pct}
focus_ath_ids <- c(63291,12150,56885,5223,184205,121309)
focus_ath <- win_pct |> 
  filter(age >= 19 & age <= 40) |>
  filter(compid %in% focus_ath_ids)

win_pct |> 
  filter(age >= 20 & age <= 40) |> 
  ggplot(data = _,aes(x = age,y = cum_win_pct,group = compid)) + 
  facet_wrap(~gender,ncol = 1) +
  geom_line(alpha = 0.1) + 
  geom_line(data = focus_ath,aes(color = name),size = 1.05) + 
  scale_color_brewer(palette = 'Set2') + 
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = 'Age',y = 'Cumulative Win Pct',
       color = NULL,caption = ) +
  ggtitle(label = '') +
  theme(legend.direction = 'horizontal',legend.position = 'bottom')
```

### Cumulative Podium Percentage
Klaebo looks just as good here, maybe even closer to the standard set by Vaelbe.

```{r,cum-pod-pct}
focus_ath_ids <- c(63291,12150,15232,57203,183629,5223,184205,121309)
focus_ath <- win_pct |> 
  filter(age >= 19 & age <= 40) |> 
  filter(compid %in% focus_ath_ids)
win_pct |> 
  filter(age >= 19 & age <= 40) |> 
  ggplot(data = _,aes(x = age,y = cum_pod_pct,group = compid)) + 
  facet_wrap(~gender,ncol = 1) +
  geom_line(alpha = 0.1) + 
  geom_line(data = focus_ath,aes(color = name),size = 1.05) + 
  #scale_color_brewer(palette = 'Set2') + 
  scale_y_continuous(labels = scales::percent_format()) +
  guides(color = guide_legend(nrow = 3)) +
  labs(x = 'Age',y = 'Cumulative Podium Pct',
       color = NULL,caption = ) +
  ggtitle(label = '') +
  theme(legend.direction = 'horizontal',legend.position = 'bottom')
```

