---
title: "Some Jan Thomas Jenssen Results"
author: "Joran Elias"
date: "2023-11-29"
categories: [world cup,ruka,jan thomas jenssen]
draft: false
---

```{r pkgs,message=FALSE}
library(tidyverse)
library(statskier2)

statskier_connect()
```

I'll be honest, I'm posting this mostly just to test some of my internal blog
setup stuff. I was poking around with Jan Thomas Jenssen's race results, given his
dramatic win last weekend and found one mostly boring thing, and one slightly
confusing thing.

```{r ath-data}
p_dst <- ath_plot_dst(ath_names = 'JENSSEN Jan Thomas',races = 'fis',by_tech = TRUE)
p_spr <- ath_plot_spr(ath_names = 'JENSSEN Jan Thomas',races = 'fis',by_tech = TRUE)
```

### Mostly Boring Thing
He's been consistently better at freestyle than classic.

```{r dst-plot}
p_dst$plot
```

### Slightly Confusing Thing
He did a sprint race where he finished 315th? And it's not shown in the plot,
but in that race his qualification time was 25th. It was a Norwegian junior
national championship race from 2015. There must be something about the format
of how they ran that sprint race that's not entirely clear from the results, 
because a single sprint race with 475 competitors seems dubious, even for Norway.

Maybe the "real" results were split up by age group somehow that isn't reflected
on the FIS website? I'm not sure. You can check the [qualification](https://www.fis-ski.com/DB/general/results.html?sectorcode=CC&raceid=25312) and
[final](https://www.fis-ski.com/DB/general/results.html?sectorcode=CC&raceid=25313) results yourself.

```{r spr-plot}
p_spr$plot
```