---
title: "Impacts of Sprint QF Heat Choice"
author: "Joran Elias"
date: "2023-12-31"
categories: [world cup,sprint,heats]
draft: false
execute: 
  message: false
---

```{r}
#| label: pkgs
#| message: false
library(MASS)
library(tidyverse)
library(statskier2)
library(RColorBrewer)
library(broom)

theme_set(theme_minimal())

statskier_connect()

v_spr_hts <- dplyr::tbl(..statskier_pg_con..,'v_sprint_heats')
```

```{r}
#| label: pull-data
maj_int_spr <- v_spr_hts |> 
  filter(primary_tag %in% c('wc','tds','owg','wsc') & 
           season >= '2015-2016') |> 
  collect()

qf_key <- maj_int_spr |> 
  filter(!is.na(qf)) |> 
  select(eventid,compid,qf) |> 
  distinct()
sf_key <- maj_int_spr |> 
  filter(!is.na(sf)) |> 
  select(eventid,compid,sf) |> 
  distinct()

ath_fn <- maj_int_spr |> 
  filter(!is.na(fn)) |> 
  select(-qf,-sf) |> 
  left_join(qf_key,by = c('eventid','compid')) |> 
  left_join(sf_key,by = c('eventid','compid'))
```

## Sprint QF Heat Choices
I was having a conversation earlier with some friends about the strategy behind
the choice of QF heat in World Cup sprint races, and realized I hadn't looked
at the actual empirical data on that in quite a while.

The background here is that athletes can choose their QF heat, and there's some
debate about how to strategize that, balancing: 

- what other competitors have already selected a particular heat
- the fact that QFs 1 & 2 are funneled into SF1, giving them an extra few minutes of
rest before the final (assuming you make it through)

So what have the athletes been doing? The following plot shows the proportion of
each QF comprising the athletes in the final in each season.

```{r}
#| label: spr-heat-qf-bar-plot
qf_pct <- ath_fn |> 
  group_by(season,gender,qf) |> 
  summarise(
    n = n(),
    .groups = 'drop'
  ) |> 
  group_by(season,gender) |> 
  mutate(pct = n / sum(n),
         qf_fct = factor(qf,levels = as.character(5:1)))

brbg <- RColorBrewer::brewer.pal(n = 5,'BrBG')
brbg[3] <- gplots::col2hex('darkgrey')

ggplot(data = qf_pct,aes(x = season,y = n,fill = qf_fct)) + 
  facet_wrap(~gender,ncol = 1) +
  geom_col(position = position_fill()) + 
  geom_vline(xintercept = 8.5,linewidth = 1.5) +
  scale_fill_manual(values = brbg) + 
  scale_y_continuous(labels = scales::label_percent()) + 
  labs(x = NULL,y = NULL,fill = 'QF',caption = 'statisticalskier.netlify.app') +
  ggtitle(
    label = "QF Choices of Sprint Finalists",
    subtitle = "Proportion of each QF represented in the top six for each season."
  ) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme(
    strip.text = element_text(face = 'bold',size = 12),
    legend.position = 'bottom',
    legend.direction = 'horizontal',
    panel.grid = element_blank()
  )
```

As you can see, there's been a modest but consistent preference for QFs 1 & 2 
for both men & women, and while it has bounced around it has remained fairly
consistent over time. (Note that I've separated the current season slightly, since
we only have a handful of sprint races so far, so we can't necessarily read much
into any changes yet.)

The problem here is that causality is almost impossible to establish. Are skiers
more likely to reach the finals because they are choosing QFs 1/2, or are the 
sprinters choosing QFs 1/2 already the best sprinters, so they'd be more likely
to reach the finals anyway?

A very crude way to try to disentangle that might be to fit a proportional odds
logistic regression (POLR) with a variable that acts as some sort of proxy for skier
strength.

Specifically, we can model a response variable of finishing place in the final 
(1st - 6th) as a function of their sprint FIS points at the time of the race 
(to control for skier strength) and the QF they originated from.

If you've never heard of POLR the basic idea is that it's fitting 5 separate 
binary logistic regressions for each response variable category transition: 6th to 5th,
5th to 4th, 4th to 3rd, etc. and estimating the impact of each variable on 
finishing in one of the higher categories.

You can see the results in the following plot, which due to the way the model 
works presents everything relative to QF1.

```{r}
#| label: polr-mdl-coefs
fis_pts <- arrow::read_parquet(file = '~/Documents/code/official-fis-points-lists/output/fis_pts_list_2023-12-28.parquet') |> 
  select(listid = Listid,
         #listname = Listname,
         listnum = ListNum,
         compid = Competitorid,
         year = year,
         calc_date = Calculationdate,
         spr_points = SPpoints) |> 
  filter(year >= 2015) |>
  distinct()

polr_data <- ath_fn |>
  filter(heat_rank <= 6) |> 
  mutate(date = as.Date(date)) |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(qf_fct = factor(qf),
         qf1 = factor(as.numeric(qf == 1)),
         qf2 = factor(as.numeric(qf == 2)),
         qf3 = factor(as.numeric(qf == 3)),
         qf4 = factor(as.numeric(qf == 4)),
         qf5 = factor(as.numeric(qf == 5)),
         fn_fct = factor(heat_rank,levels = as.character(6:1)),
         spr_points_scl = (spr_points - mean(spr_points,na.rm = TRUE)) / sd(spr_points,na.rm = TRUE))

m_men <- MASS::polr(fn_fct ~ spr_points_scl + qf_fct,
                    data = polr_data,
                    subset = polr_data$gender == 'Men')
m_wom <- MASS::polr(fn_fct ~ spr_points_scl + qf_fct,
                    data = polr_data,
                    subset = polr_data$gender == 'Women')

men_coef <- broom::tidy(m_men,exponentiate = TRUE,conf.int = TRUE)
wom_coef <- broom::tidy(m_wom,exponentiate = TRUE,conf.int = TRUE)

bind_rows(
  men_coef |> mutate(gender = 'Men'),
  wom_coef |> mutate(gender = 'Women')
) |> 
  filter(grepl('^qf',term)) |> 
  mutate(term_lbl = gsub('qf_fct','QF',term)) |> 
  select(gender,term_lbl,estimate,starts_with('conf')) |> 
  ggplot(data = _,aes(x = term_lbl)) + 
  facet_wrap(~gender,ncol = 1,scales = 'free_y') +
  geom_pointrange(aes(y = estimate,ymin = conf.low,ymax = conf.high),size = 0.2) +
  labs(x = NULL,
       y = 'Relative Likelihood',
       caption = 'statisticalskier.netlify.app') + 
  ggtitle(
    label = 'Relative likelihood of a better result in sprint final compared to QF1',
    subtitle = 'e.g. men from QF3 are ~0.2x less likely to finish\n1st vs 2nd, 2nd vs 3rd, 3rd vs 4th, etc compared to QF1'
  ) +
  theme(
    strip.text = element_text(face = 'bold',size = 12)
  )

```

It's tempting to think we've solved the causality direction problem here since these
results "controlled for" skier strength, but I don't really buy it. This is still
operating on fundamentally observational data; ideally we'd run a bunch of sprint 
races with athletes randomly assigned to QFs, but for some reason FIS hasn't
taken me up on that idea.

Also, I think that FIS sprint points are a fairly noisy measure of "skier strength"
since it's really just measuring qualification speed. That's almost certainly
*sort of* correlated with overall results in the heats, but I think most World
Cup athletes would tell you that succeeding the in heats is a pretty different
sort of skill than succeeding in qualification.

