---
title: "The 20k Freestyle Race in Goms Was Absurdly Fast"
author: "Joran Elias"
date: "2024-01-30"
categories: [world cup,distance,freestyle,mass start,goms,pace]
draft: false
---

```{r pkgs,message=FALSE}
library(tidyverse)
library(statskier2)
library(ggrepel)
library(hms)
library(gt)

theme_set(theme_minimal())
statskier_connect()
```

```{r}
ev_ids <- c(19590,19591)

v_dst_maj_int <- dplyr::tbl(..statskier_pg_con..,"v_distance_maj_int")
v_dst_splits <- dplyr::tbl(..statskier_pg_con..,"v_distance_splits")

goms_splits <- 
  v_dst_splits |> 
  filter(eventid %in% ev_ids) |> 
  collect()

dst_res <- 
  v_dst_maj_int |> 
  filter(between(length,5,50) & !is.na(rank) & rank <= 10 & format %in% c("Interval","Mass") & tech == "F") |> 
  collect()
```

```{r}
dst_res <- 
  dst_res |> 
  mutate(pace = time / length)

avg_pace <- 
  dst_res |> 
  group_by(season,date,eventid,gender,location,site,primary_tag,length,format) |> 
  summarise(
    avg_sec_per_km = mean(pace,na.rm = TRUE),
    winning_time = min(time),
    .groups = "drop")
goms_avg_pace <- 
  avg_pace |> 
  filter(date == "2024-01-28") |> 
  mutate(lbl = "Goms")
tds_climb <- 
  data.frame(
    date = c("2024-01-01","2024-01-01"),
    avg_sec_per_km = c(210,235),
    gender = c("Men","Women"),
    lbl = rep("TdS Alpe Cermis",2)
  )
```

A reader emailed me to comment on the blistering pace of the recent 20k freestyle 
mass start races in Goms. And indeed, they were quite fast!

```{r}
avg_pace |> 
  ggplot(data = _,aes(x = as.Date(date),y = avg_sec_per_km,color = format)) + 
  facet_wrap(~gender,ncol = 2) +
  geom_point(alpha = 0.75) + 
  geom_smooth(formula = y~x,method = MASS::rlm,se = FALSE,linewidth = 0.8) +
  geom_point(data = goms_avg_pace,color = "orange",size = 3) + 
  geom_text_repel(data = goms_avg_pace,aes(label = lbl),color = "black",nudge_y = -15,size = 3) +
  geom_text_repel(data = tds_climb,aes(label = lbl),color = "black",size = 3) +
  scale_color_brewer(palette = "Set2") +
  labs(x = NULL,y = "Seconds / Kilometer",color = NULL) + 
  ggtitle(
    label = "Average freestyle pace of top ten skiers in WC/TdS/WSC/OWG events",
    subtitle = "Only freestyle interval & mass start events (excluding skiathlons) between 5-50km") +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  )
```

Everything I've heard suggests that the course was very icy and fast, so that was
surely a contributing factor. It is a **little** strange maybe to have such a fast
pace in a mass start these days, but it's not unprecedented.

The men's race was the fastest (by this measure, average pace of the top ten) that
I have on record, but just barely. The women's was only the second fastest. I'd
certainly call them mild outliers for mass start races, but if you squint maybe 
within the realm of possibility given the "normal" variation in the scatterplot
above.

```{r}
avg_pace |> 
  filter(gender == "Men") |> 
  mutate(winning_time_lbl = hms(winning_time)) |> 
  arrange(avg_sec_per_km) |> 
  select(date,primary_tag,location,site,length,format,avg_sec_per_km,winning_time_lbl) |> 
  slice_min(order_by = avg_sec_per_km,n = 10) |> 
  gt() |> 
  fmt_number(columns = avg_sec_per_km) |> 
  cols_label(
    date = "Date",
    primary_tag = "Event",
    location = "Nation",
    site = "Site",
    length = "Km",
    format = "Format",
    avg_sec_per_km = "Avg Leader Pace",
    winning_time_lbl = "Winning Time"
  ) |> 
  tab_header(
    title = "Fastest Men's Freestyle Races"
  ) |> 
  tab_style(style = cell_text(weight = "bold"),locations = list(cells_column_labels(),cells_title())) |> 
  tab_style(style = cell_text(align = "left"),locations = list(cells_column_labels(columns = 1),cells_title()))
```

```{r}
avg_pace |> 
  filter(gender == "Women") |> 
  mutate(winning_time_lbl = hms(winning_time)) |> 
  arrange(avg_sec_per_km) |> 
  select(date,primary_tag,location,site,length,format,avg_sec_per_km,winning_time_lbl) |> 
  slice_min(order_by = avg_sec_per_km,n = 10) |> 
  gt() |> 
  fmt_number(columns = avg_sec_per_km) |> 
  cols_label(
    date = "Date",
    primary_tag = "Event",
    location = "Nation",
    site = "Site",
    length = "Km",
    format = "Format",
    avg_sec_per_km = "Avg Leader Pace",
    winning_time_lbl = "Winning Time"
  ) |> 
  tab_header(
    title = "Fastest Women's Freestyle Races"
  ) |> 
  tab_style(style = cell_text(weight = "bold"),locations = list(cells_column_labels(),cells_title())) |> 
  tab_style(style = cell_text(align = "left"),locations = list(cells_column_labels(columns = 1),cells_title()))
```