---
title: "Podium Counts & Percents"
author: "Joran Elias"
date: "2024-02-14"
categories: [world cup,sprint,distance]
draft: false
fig-width: 9
fig-height: 9
---

```{r setup}
library(tidyverse)
library(statskier2)
library(ggridges)
library(ggrepel)
library(ggforce)
library(patchwork)

statskier_connect()

theme_set(theme_minimal())

v_dst_maj_int <- dplyr::tbl(..statskier_pg_con..,"v_distance_maj_int")
v_spr_maj_int <- dplyr::tbl(..statskier_pg_con..,"v_sprint_maj_int")
dst_maj_int <- v_dst_maj_int |> collect()
spr_maj_int <- v_spr_maj_int |> collect()
```

Devon Kershaw was nerding out a bit on his podcast this week about both how many
podium Klæbo has and the percentage of his races that are podiums, so I thought 
I'd make some graphs.

Part of the reason this caught my ear was that after a recent transcription push
I now have a close enough to complete record of all major international (World Cup,
Tour de Ski, Olympic & World Championship) that I actually feel comfortable making
summaries that stretch well back into the 80's.

It's not 100% complete. I know that I'm missing 1 World Cup race for sure, and 
another where the results I have are incomplete, but it's good enough for some
high level stuff with caveats.

One of those caveats is that I count things differently than FIS, so don't
expect these numbers to exactly match what FIS publishes. FIS will include relays
in counts of podiums, and stage race overall results, which I prefer not to. I also 
prefer to exclude pursuit starts where World Cup results were based on the "time of the day".

However, I can include reasonably complete data for elite skiers from the 80's now.

```{r data}
dst_pods <- dst_maj_int |> 
  filter(!is.na(rank) & 
           format != "Pursuit" & 
           season >= "1981-1982") |> 
  summarise(
    n_pods = sum(rank <= 3),
    n_wins = sum(rank == 1),
    n = n_distinct(eventid),
    pct_pod = mean(rank <= 3),
    pct_win = mean(rank == 1),
    .by = c('gender','compid','fisid','name')
  ) |> 
  arrange(desc(pct_pod)) |> 
  mutate(type = 'Distance',
         name_lbl = statskier2::extract_all_caps(name),
         name_lbl = stringr::str_to_title(name_lbl)) |> 
  mutate(fill_highlight = 
           case_when(
             name_lbl %in% c("Svan","Vaelbe","Bolshunov","Daehlie","Johaug","Bjoergen") ~ "Tier 1",
             name_lbl %in% c("Smirnov","Lazutina","Kowalczyk","Belmondo") ~ "Tier 2",
             name_lbl %in% c("Karlsson","Klaebo","Pettersen","Andersson","Kirvesniemi","Matikanen Kallstroem","Elofsson") ~ "Tier 3",
             name_lbl %in% c("Skari Martinsen","Egorova","Sundby","Weng","Smigun-Vaehi","Neumannova","Cologna","Kalla","Northug") ~ "Tier 4",
             TRUE ~ "Other"
           ))

spr_pods <- spr_maj_int |> 
  filter(!is.na(rank)) |> 
  summarise(
    n_pods = sum(rank <= 3),
    n_wins = sum(rank == 1),
    n = n_distinct(eventid),
    pct_pod = mean(rank <= 3),
    pct_win = mean(rank == 1),
    .by = c('gender','compid','fisid','name')
  ) |> 
  arrange(desc(pct_pod)) |> 
  mutate(type = 'Sprint',
         name_lbl = statskier2::extract_all_caps(name),
         name_lbl = stringr::str_to_title(name_lbl)) |> 
  mutate(fill_highlight = 
           case_when(
             name_lbl %in% c("Klaebo") ~ "Tier 1",
             name_lbl %in% c("Falla","Bjoergen") ~ "Tier 2",
             name_lbl %in% c("Skari Martinsen","Svahn") ~ "Tier 3",
             name_lbl %in% c("Nilsson","Sundling","Hetland","Majdic","Pellegrino","Northug","Dahlqvist") ~ "Tier 4",
             TRUE ~ "Other"
           ))

comb_pods <- 
  bind_rows(
    dst_maj_int |> 
      filter(!is.na(rank) & 
               format != "Pursuit" & 
               season >= "1981-1982") |> 
      select(eventid,gender,compid,fisid,name,rank),
    spr_maj_int |> 
      filter(!is.na(rank)) |> 
      select(eventid,gender,compid,fisid,name,rank)
  ) |> 
  summarise(
    n_pods = sum(rank <= 3),
    n_wins = sum(rank == 1),
    n = n_distinct(eventid),
    pct_pod = mean(rank <= 3),
    pct_win = mean(rank == 1),
    .by = c('gender','compid','fisid','name')
  ) |> 
  arrange(desc(pct_pod)) |> 
  mutate(type = 'Overall',
         name_lbl = statskier2::extract_all_caps(name),
         name_lbl = stringr::str_to_title(name_lbl)) |> 
  mutate(fill_highlight = 
           case_when(
             name_lbl %in% c("Svan","Vaelbe","Klaebo","Daehlie","Johaug","Bjoergen") ~ "Tier 1",
             name_lbl %in% c("Bolshunov","Smirnov","Skari Martinsen","Lazutina","Falla","Belmondo","Northug","Kowalczyk") ~ "Tier 2",
             TRUE ~ "Other"
           ))
```

```{r plots-setup}
N <- 15
P <- 0.15
x_theme <- theme(
  plot.title.position = "plot",
  plot.title = element_text(face = "bold"),
  legend.position = "bottom",
  legend.direction = "horizontal",
  plot.caption = element_text(face = "italic")
)
lbls <- labs(
  x = "Number of podiums",
  y = "Percent podiums",
  color = NULL,
  fill = NULL,
  caption = "WC/TdS/OWG/WSC; pursuit starts & relays excluded; DNF/DSQs not counted against % podium\nstatisticalskier.netlify.app")
```

### Overall Podiums
Starting with distance & sprint combined, which looks like this:
```{r plot-overall}
comb_pods |> 
  filter(n_pods > N & pct_pod > P) |> 
  ggplot(data = _,aes(x = n_pods,y = pct_pod)) + 
  geom_mark_hull(
    aes(fill = fill_highlight,filter = fill_highlight != "Other"),
    color = NA,alpha = 0.25,expand = unit(3,"mm")) +
  geom_point() +
  geom_text_repel(aes(label = name_lbl,color = gender),size = 3,fontface = "bold") + 
  scale_y_continuous(labels = scales::label_percent()) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set2",guide = "none") +
  lbls + 
  ggtitle(
    label = "Overall: percent vs number of podiums",
    subtitle = glue::glue("Skiers with a minimum of {n} podiums & {p}% podiums.",n = N,p = P*100)
  ) + 
  x_theme
```
In this case, there's Bjœrgen, Klæbo, Johaug, Dæhlie, Vaelbe, Svan and then 
there's maybe a second group ranging from Bolshunov to Kowalczyk. 

### Distance Podiums
Focusing on distance only yields basically the same picture at the top, except we've swapped
Bolshunov in for Klæbo. Then there are maybe three small groups below that elite tier.

```{r plot-distance}
dst_pods |> 
  filter(n_pods > N & pct_pod > P) |> 
  ggplot(data = _,aes(x = n_pods,y = pct_pod)) + 
  geom_mark_hull(
    aes(fill = fill_highlight,filter = fill_highlight != "Other"),
    color = NA,alpha = 0.25,expand = unit(3,"mm")) +
  geom_point() +
  geom_text_repel(aes(label = name_lbl,color = gender),size = 3,fontface = "bold") + 
  scale_y_continuous(labels = scales::label_percent()) + 
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set2",guide = "none") +
  lbls + 
  ggtitle(
    label = "Distance: percent vs number of podiums",
    subtitle = glue::glue("Skiers with a minimum of {n} podiums & {p}% podiums. (WC/TdS/OWG/WSC; pursuit starts & relays excluded)",n = N,p = P*100)
  ) + 
  x_theme
```
### Sprint Podiums
Lastly, focusing on only sprints we see that Klæbo really is off all on his own.
The only person remotely close in terms of percentages is Bente Skari Martinsen
who raced in an entirely different era of sprinting. Bjœrgen & Fall are off on
their own over to the right and then there's group in the center around Hetland & 
Majdic.
```{r plot-sprint}
spr_pods |> 
  filter(n_pods > N & pct_pod > P) |> 
  ggplot(data = _,aes(x = n_pods,y = pct_pod)) +
  geom_mark_ellipse(
    aes(fill = fill_highlight,filter = fill_highlight != "Other"),
    color = NA,alpha = 0.25,expand = unit(3,"mm")) +
  geom_point() +
  geom_text_repel(aes(label = name_lbl,color = gender),size = 3,fontface = "bold") + 
  scale_y_continuous(labels = scales::label_percent()) + 
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set2",guide = "none") +
  lbls + 
  ggtitle(
    label = "Sprint: percent vs number of podiums",
    subtitle = glue::glue("Skiers with a minimum of {n} podiums & {p}% podiums. (WC/TdS/OWG/WSC; relays excluded)",n = N,p = P*100)
  ) + 
  x_theme
```