---
title: "Trondheim 10k Classic Predictions"
author: "Joran Elias"
date: "2023-12-16"
categories: [world cup,distance,classic,interval start,predictions,trondheim]
draft: false
---

```{r pkgs,message=FALSE}
library(tidyverse)
library(statskier2)
library(gt)
library(gtExtras)
library(ggpubr)

theme_set(theme_bw())
```

### Predictions
Haha...I wonder how much I'd have to put my thumb on the scale for this page 
rank method to stop picking Friday to win? I mean, if she does bad enough for 
long enough it will catch up eventually.

```{r pred-data}
location <- "Trondheim"
len <- "10k"
tech <- "Classic"
mens_preds <- 
  readr::read_csv(file = '2023-12-17_men_cl_dst_trondheim.csv',
                  show_col_types = FALSE)
womens_preds <- 
  readr::read_csv(file = '2023-12-17_wom_cl_dst_trondheim.csv',
                  show_col_types = FALSE)
```

#### Men's 10k Classic Predictions
```{r tbl-men}
mens_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = glue::glue("{loc} {len} {tech}",loc = location,len = len,tech = tech),
             subtitle = "Men") |> 
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank) 
```

#### Women's 10k Classic Predictions
```{r tbl-wom}
womens_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = glue::glue("{loc} {len} {tech}",loc = location,len = len,tech = tech),
             subtitle = "Women") |>
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank)
```

### Update
```{r db-con}
statskier_connect()
v_dst_maj_int <- dplyr::tbl(..statskier_pg_con..,'v_distance_maj_int')
```

```{r race-data}
dst_res <- v_dst_maj_int |> 
  filter(eventid %in% c(19278,19279)) |> 
  collect()

fis_pts <- readRDS(file = '~/Documents/code/official-fis-points-lists/output/fis_pts_list.rds') |> 
  select(listid = Listid,
         listname = Listname,
         listnum = ListNum,
         compid = Competitorid,
         year = year,
         calc_date = Calculationdate,
         dst_points = DIpoints) |> 
  filter(year >= 2020)

men_comb <- dst_res |> 
  filter(gender == 'Men') |> 
  select(date,compid,rank) |> 
  mutate(date = as.Date(date)) |> 
  full_join(mens_preds,by = 'compid') |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(
    fis_pts_rank = rank(dst_points,ties.method = 'min')) |> 
  filter(!is.na(rank)) |> 
  arrange(rank)

wom_comb <- dst_res |> 
  filter(gender == 'Women') |> 
  select(date,compid,rank) |> 
  mutate(date = as.Date(date)) |>
  full_join(womens_preds,by = 'compid') |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(
    fis_pts_rank = rank(dst_points,ties.method = 'min')) |> 
  filter(!is.na(rank)) |> 
  arrange(rank)
```

You know, I think the start lists I'm using when generating the predictions are
as accurate as they could be, since Frida isn't on the PDF start list now. Hm.

```{r men-perf}
men_comb |> 
  select(compid,rank,page_rank_pos,fis_pts_rank) |> 
  pivot_longer(cols = c(page_rank_pos,fis_pts_rank),names_to = 'grp',values_to = 'val') |> 
  mutate(grp = if_else(grp == 'fis_pts_rank','FIS Dst Pts List','PR Method')) |> 
  ggplot(data = _,aes(x = rank,y = val)) + 
  facet_wrap(~grp,ncol = 1) + 
  geom_point() + 
  geom_abline(slope = 1,intercept = 0,color = 'blue') + 
  stat_cor(aes(label = after_stat(..r.label..))) +
  ggtitle(label = glue::glue("Men's {loc} {len} {tech} Predicted vs True Athlete Finishing Order",
                             loc = location,len = len,tech = tech),
          subtitle = "Athletes ordered by most recent FIS Dst pts list (top) & my PR method (bottom)") +
  labs(x = 'True Finishing Place',y = 'Predicted Finishing Place')
```

```{r wom-perf}
wom_comb |> 
  select(compid,rank,page_rank_pos,fis_pts_rank) |> 
  pivot_longer(cols = c(page_rank_pos,fis_pts_rank),names_to = 'grp',values_to = 'val') |> 
  mutate(grp = if_else(grp == 'fis_pts_rank','FIS Dst Pts List','PR Method')) |> 
  ggplot(data = _,aes(x = rank,y = val)) + 
  facet_wrap(~grp,ncol = 1) + 
  geom_point() + 
  geom_abline(slope = 1,intercept = 0,color = 'blue') + 
  stat_cor(aes(label = after_stat(..r.label..))) +
  ggtitle(label = glue::glue("Women's {loc} {len} {tech} Predicted vs True Athlete Finishing Order",
                             loc = location,len = len,tech = tech),
          subtitle = "Athletes ordered by most recent FIS Dst pts list (top) & my PR method (bottom)") +
  labs(x = 'True Finishing Place',y = 'Predicted Finishing Place')
```