---
title: "Gällivare 10k Freestyle Interval Start Predictions"
author: "Joran Elias"
date: "2023-12-01"
pubDate: "Fri, 01 Dec 2023 15:35:23 MST"
categories: [world cup,distance,freestyle,interval start,predictions]
draft: false
---

```{r pkgs,message=FALSE}
library(tidyverse)
library(statskier2)
library(gt)
library(gtExtras)
library(ggpubr)

theme_set(theme_bw())
```

### Predictions
On to the second weekend in Gällivare, Sweden. I've decided I'm not going to do
any significant tinkering with the model generating these predictions until we're
between periods. Just too busy and need some longer time gaps without races to
experiment.

```{r pred-data}
mens_preds <- 
  readr::read_csv(file = '2023-12-02_men_fr_dst_gaellivare.csv',
                  show_col_types = FALSE)
womens_preds <- 
  readr::read_csv(file = '2023-12-02_women_fr_dst_gaellivare.csv',
                  show_col_types = FALSE)
```

#### Men's 10k Freestyle Predictions
```{r tbl-men}
mens_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = "Ruka 20k Freestyle Mass Start",subtitle = "Men") |> 
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank) 
```

#### Women's 10k Freestyle Predictions
I think maybe these are a little too conservative on Rosie Brennan. We'll see.
```{r tbl-wom}
womens_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = "Gällivare 10k Freestyle Interval Start",subtitle = "Women") |>
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank)
```

### Update
```{r db-con}
statskier_connect()
v_dst_maj_int <- dplyr::tbl(..statskier_pg_con..,'v_distance_maj_int')
```

```{r race-data}
dst_res <- v_dst_maj_int |> 
  filter(eventid %in% c(19200,19201)) |> 
  collect()

fis_pts <- readRDS(file = '~/Documents/code/official-fis-points-lists/output/fis_pts_list.rds') |> 
  select(listid = Listid,
         listname = Listname,
         listnum = ListNum,
         compid = Competitorid,
         year = year,
         calc_date = Calculationdate,
         dst_points = DIpoints) |> 
  filter(year >= 2020)

men_comb <- dst_res |> 
  filter(gender == 'Men') |> 
  select(date,compid,rank) |> 
  mutate(date = as.Date(date)) |> 
  full_join(mens_preds,by = 'compid') |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(
    fis_pts_rank = rank(dst_points,ties.method = 'min')) |> 
  arrange(rank)

wom_comb <- dst_res |> 
  filter(gender == 'Women') |> 
  select(date,compid,rank) |> 
  mutate(date = as.Date(date)) |>
  full_join(womens_preds,by = 'compid') |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(
    fis_pts_rank = rank(dst_points,ties.method = 'min')) |> 
  arrange(rank)
```

My method did modestly better for both men & women, but not by a ton:

```{r men-perf}
men_comb |> 
  select(compid,rank,page_rank_pos,fis_pts_rank) |> 
  pivot_longer(cols = c(page_rank_pos,fis_pts_rank),names_to = 'grp',values_to = 'val') |> 
  mutate(grp = if_else(grp == 'fis_pts_rank','FIS Dst Pts List','PR Method')) |> 
  ggplot(data = _,aes(x = rank,y = val)) + 
  facet_wrap(~grp,ncol = 1) + 
  geom_point() + 
  geom_abline(slope = 1,intercept = 0,color = 'blue') + 
  stat_cor(aes(label = after_stat(..r.label..))) +
  ggtitle(label = "Men's Gällivare 10k Freestyle Interval Predicted vs True Athlete Finishing Order",
          subtitle = "Athletes ordered by most recent FIS Dst pts list (top) & my PR method (bottom)") +
  labs(x = 'True Finishing Place',y = 'Predicted Finishing Place')
```

```{r wom-perf}
wom_comb |> 
  select(compid,rank,page_rank_pos,fis_pts_rank) |> 
  pivot_longer(cols = c(page_rank_pos,fis_pts_rank),names_to = 'grp',values_to = 'val') |> 
  mutate(grp = if_else(grp == 'fis_pts_rank','FIS Dst Pts List','PR Method')) |> 
  ggplot(data = _,aes(x = rank,y = val)) + 
  facet_wrap(~grp,ncol = 1) + 
  geom_point() + 
  geom_abline(slope = 1,intercept = 0,color = 'blue') + 
  stat_cor(aes(label = after_stat(..r.label..))) +
  ggtitle(label = "Women's Gällivare 10k Freestyle Interval Predicted vs True Athlete Finishing Order",
          subtitle = "Athletes ordered by most recent FIS Dst pts list (top) & my PR method (bottom)") +
  labs(x = 'True Finishing Place',y = 'Predicted Finishing Place')
```