---
title: "World Cup Winning Streaks By Nation"
author: "Joran Elias"
date: "2023-12-17"
categories: [world cup,distance,sprint,winning streak]
draft: false
execute: 
  message: false
---

```{r}
#| label: pkgs
#| message: false
library(tidyverse)
library(statskier2)
library(gt)

theme_set(theme_minimal())

statskier_connect()

v_dst_tbl <- dplyr::tbl(..statskier_pg_con..,'v_distance_maj_int')
v_spr_tbl <- dplyr::tbl(..statskier_pg_con..,'v_sprint_maj_int')
```

```{r}
#| label: pull-data
dst <- v_dst_tbl |> 
  filter(season >= '1991-1992' & 
           !is.na(rank) & 
           rank == 1) |> 
  collect() |> 
  arrange(gender,date)
spr <- v_spr_tbl |> 
  filter(season >= '2006-2007' & 
           !is.na(rank) & 
           rank == 1) |> 
  collect() |> 
  arrange(gender,date)
```

```{r}
build_ath_counts <- function(nms){
  nms <- extract_all_caps(nms)
  nms <- str_to_title(nms)
  nms_tbl <- sort(table(nms),decreasing = TRUE)
  paste(paste(names(nms_tbl),paste0("(",nms_tbl,")")),collapse = "<br>")
}
```

```{r}
#| label: process-data
dst_streaks <- 
  dst |> 
  group_by(gender) |> 
  mutate(streak_id = consecutive_id(nation)) |> 
  group_by(gender,streak_id) |> 
  summarise(nation = nation[1],
            streak_start = min(date),
            streak_end = max(date),
            streak_len_days = as.integer(as.Date(streak_end) - as.Date(streak_start)) + 1,
            streak_len_n = n(),
            n_skiers = n_distinct(compid),
            skier_names = build_ath_counts(name),
            .groups = 'drop') |> 
  mutate(streak_end = if_else(streak_end == '2023-12-17','?',streak_end))

spr_streaks <- 
  spr |> 
  group_by(gender) |> 
  mutate(streak_id = consecutive_id(nation)) |> 
  group_by(gender,streak_id) |> 
  summarise(nation = nation[1],
            streak_start = min(date),
            streak_end = max(date),
            streak_len_days = as.integer(as.Date(streak_end) - as.Date(streak_start)) + 1,
            streak_len_n = n(),
            n_skiers = n_distinct(compid),
            skier_names = build_ath_counts(name),
            .groups = 'drop') |> 
  mutate(streak_end = if_else(streak_end == '2023-12-17','?',streak_end))
```

## National Winning Streaks
The Norwegian men are currently on a 17 race winning streak in distance races which
frankly doesn't seem likely to end anytime soon.

In that light, I thought I'd catalogue the longest such streaks I can find. This
first table lists the largest distance winning streaks. As you can see, the 
Norwegian men are on the cusp of besting the Norwegian women's streak of 17 
spanning 2014 & 2015.

```{r}
#| label: dst-streak-tbls
dst_streaks |> 
  group_by(gender) |> 
  slice_max(order_by = streak_len_n,n = 5) |> 
  select(-streak_id,-streak_len_days,-n_skiers) |> 
  gt(row_group_as_column = TRUE) |> 
  tab_options(table.font.size = pct(80)) |> 
  cols_label(
    nation = "Nation",
    streak_start = "Start",
    streak_end = "End",
    streak_len_n = "# Races",
    skier_names = "Athletes"
  ) |> 
  fmt_markdown(columns = skier_names) |> 
  tab_style(
    style = cell_text(v_align = 'top'),
    locations = cells_body(columns = -c(skier_names))
  ) |>
  tab_style(
    style = cell_text(align = 'right'),
    locations = cells_body(columns = skier_names)
  ) |> 
  tab_style(
    style = cell_text(weight = 'bold'),
    locations = list(cells_column_labels(),cells_row_groups())
  ) |> 
  tab_header(title = "National Winning Streaks: Distance",
             subtitle = "Includes World Cup, OWG, WSC & TdS events from 1991") |> 
  tab_style(
    style = cell_text(align = 'left'),
    locations = cells_title()
  ) |> 
  tab_footnote(footnote = "This streak may have extended earlier but my data is incomplete pre-1991.",
               locations = cells_body(rows = streak_start == '1991-12-08',columns = streak_start),placement = 'right')
```

Sprinting winning streaks by a nation haven't been as long, topping out at 9.

```{r}
#| label: spr-streak-tbls
spr_streaks |> 
  group_by(gender) |> 
  slice_max(order_by = streak_len_n,n = 5) |> 
  select(-streak_id,-streak_len_days,-n_skiers) |> 
  gt(row_group_as_column = TRUE) |> 
  tab_options(table.font.size = pct(80)) |> 
  cols_label(
    nation = "Nation",
    streak_start = "Start",
    streak_end = "End",
    streak_len_n = "# Races",
    skier_names = "Athletes"
  ) |> 
  fmt_markdown(columns = skier_names) |> 
  tab_style(
    style = cell_text(v_align = 'top'),
    locations = cells_body(columns = -c(skier_names))
  ) |>
  tab_style(
    style = cell_text(align = 'right'),
    locations = cells_body(columns = skier_names)
  ) |> 
  tab_style(
    style = cell_text(weight = 'bold'),
    locations = list(cells_column_labels(),cells_row_groups())
  ) |> 
  tab_header(title = "National Winning Streaks: Sprint",
             subtitle = "Includes World Cup, OWG, WSC & TdS events from 2006") |> 
  tab_style(
    style = cell_text(align = 'left'),
    locations = cells_title()
  )
```

