---
title: "The Milan Olympic Sprint Course Was Amazingly Long"
author: "Joran Elias"
date: "2026-02-12"
categories: [world cup,sprint,qualification,olympics]
draft: false
---

```{r pkgs,message=FALSE}
library(tidyverse)
library(ggokabeito)
library(statskier2)
library(slider)
library(qgam)

library(geomtextpath)
library(ggbeeswarm)
library(ggrepel)

statskier_connect()

theme_set(
  theme_minimal() + 
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      plot.title.position = "plot",
      strip.text = element_text(hjust = 0,size = rel(1.1)),
      legend.position = "bottom",
      legend.direction = "horizontal"
    )
)
```

```{r}
v_spr_maj_int <- dplyr::tbl(..statskier_pg_con..,"v_sprint_maj_int")
v_spr_heats <- dplyr::tbl(..statskier_pg_con..,"v_sprint_heats")
```

```{r}
spr_qual_times <- 
  v_spr_maj_int |> 
  filter(!is.na(rankqual) & !is.na(time)) |> 
  collect() |> 
  filter(rankqual <= 30)

spr_heat_times <- 
  v_spr_heats |> 
  collect() |> 
  filter(eventid %in% spr_qual_times$eventid)
```

```{r}
spr_qual_best <- 
  spr_qual_times |> 
  slice_min(
    order_by = time,
    n = 1,
    with_ties = FALSE,
    by = c(eventid,tech,gender,date)
  )

spr_heat_best <- 
  spr_heat_times |> 
  mutate(rnd = substr(heat,1,1)) |> 
  slice_min(
    order_by = heat_time,
    n = 1,
    with_ties = FALSE,
    by = c(eventid,tech,gender,date,rnd)
  ) |> 
  mutate(
    time = heat_time
  )
```

```{r}
common_events <- 
  inner_join(
    spr_qual_best,
    spr_heat_best,
    by = "eventid"
  ) |> 
  select(eventid) |> 
  distinct()
```

```{r}
time_prog <- 
  bind_rows(
    spr_qual_best |> 
      select(eventid,tech,gender,date,time) |> 
      mutate(rnd = "qual") |> 
      semi_join(common_events,by = "eventid"),
    spr_heat_best |> 
      select(eventid,tech,gender,date,rnd,time) |> 
      semi_join(common_events,by = "eventid")
  ) |> 
  mutate(
    rnd_lbl = 
      case_when(
        rnd == "qual" ~ "Qual",
        rnd == "1" ~ "QF",
        rnd == "2" ~ "SF",
        rnd == "3" ~ "FN"
      ),
    rnd_lbl = factor(rnd_lbl,levels = c("Qual","QF","SF","FN"),ordered = TRUE)
  ) |> 
  arrange(date,eventid,rnd_lbl) |> 
  mutate(
    milan_owg = if_else(date == as.Date("2026-02-10"),"y","n")
  ) |> 
  mutate(
    has_all_rnds = all(!is.na(time) & all(time > 100)),
    .by = eventid
  ) |> 
  filter(has_all_rnds)
```

```{r}
rnd_qs <- 
  time_prog |> 
  summarise(
    q05 = quantile(time,0.05,na.rm = TRUE),
    q50 = quantile(time,0.5,na.rm = TRUE),
    q95 = quantile(time,0.95,na.rm = TRUE),
    .by = c(gender,tech,rnd_lbl)
  ) |> 
  pivot_longer(
    cols = starts_with("q"),
    names_to = "q",
    values_to = "time"
  ) |> 
  mutate(
    q_lbl = 
      case_when(
        q == "q05" ~ "5%",
        q == "q50" ~ "50%",
        q == "q95" ~ "95%"
      )
  )
```

I've seen some discussion around how long this year's Olympic sprint course was.

This makes a difference, because physiological differences between performing for
2.5 minutes is enough different from, say 4 minutes that it really changes which
athletes are best suited to it, and demands some different training and preparation.

So let's quantify how unusually long this Olympic sprint course actually is. We
could compare the stated length, but that doesn't take into account the terrain,
which also impacts how long of an effort it takes to complete.

So instead, let's take each WC, WSC, OWG or TdS sprint race since ~2013 (that's
approximately how far back I have heat times for) and extract that fastest time
from each round. That means for each sprint race, we get four times, one each for
the qualification, QFs, SFs and the final.

The following plot shows how the Milan Olympic course stacks up against the distribution
of times across all other classic sprints. The 5%, 50% & 95% lines indicate the
5th, 50th (median) and 95% percentile times across all races in that panel.

```{r}
ggplot(
  data = rnd_qs,
  aes(x = rnd_lbl,y = time,group = q)
) + 
  facet_grid(gender~tech) + 
  geom_textpath(
    aes(label = q_lbl),
    color = "grey50",
    size = 8 / .pt,
    linetype = "dashed"
  ) + 
  geom_point(color = "grey50",size = 1) +
  geom_textline(
    data = time_prog |> 
      filter(milan_owg == "y"),
    aes(x = rnd_lbl,y = time,group = eventid),
    label = "Milan OWG",
    size = 8 / .pt,
    color = "blue",
    inherit.aes = FALSE
  ) +
  geom_point(
    data = time_prog |> 
      filter(milan_owg == "y"),
    aes(x = rnd_lbl,y = time,group = eventid),
    color = "blue",
    inherit.aes = FALSE
  ) +
  labs(
    x = "Round",
    y = "Fastest time (seconds)",
    title = "Milan Olympic Sprint Course Is Historical Long",
    subtitle = "Fastest time from each round compared to the historical (2013-2026) WC, OWG, WSC & TdS\nsprint events.",
    caption = "statisticalskier.netlify.app"
  ) + 
  scale_y_continuous(
    expand = expansion(mult = c(0.02,0.02))
  ) + 
  scale_x_discrete(
    expand = expansion(add = c(0.1,0.1))
  ) + 
  theme(
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(color = "grey",face = "italic")
  )
```

Clearly, this Olympic sprint course skied much, much longer than virtually any
other sprint course any of these athletes have had to compete on for probably 
their entire careers.

You can also see that the times got noticeably slower with each round as well. Let's
compare that as well to historical trends.

As a simple and crude way to do that, let's take each of these sets of four times
and calculate the percentage change from each heat round relative to the qualification
time, and average these percent changes.

The following plot shows the distribution of the average percent change relative to the fastest
qualification time across all sprint races since 2013. Once again, this Milan Olympic
course slowed down during the day by an historic amount.

```{r}
rnd_slopes <- 
  time_prog |> 
  arrange(eventid,rnd_lbl) |> 
  select(eventid,rnd_lbl,time) |> 
  tidyr::nest(times = -c(eventid)) |> 
  mutate(
    slope = purrr::map_dbl(
      .x = times,
      .f = \(x) with(x,coef(lm(time~rnd_lbl))[2])
    )
  ) |> 
  select(-times)

rnd_pct_chng <- 
  time_prog |> 
  arrange(eventid,rnd_lbl) |> 
  summarise(
    avg_pct_chng = mean((time - time[1]) / time[1]),
    .by = eventid
  )
```

```{r}
milan_anno <- 
  time_prog |> 
      select(eventid,gender,tech) |> 
      distinct() |> 
      left_join(
        rnd_pct_chng,
        by = "eventid"
      ) |> 
      semi_join(time_prog |> filter(milan_owg == "y"),by = "eventid") |> 
  mutate(lbl = "Milan OWG")

time_prog |> 
  select(eventid,gender,tech) |> 
  distinct() |> 
  left_join(
    rnd_pct_chng,
    by = "eventid"
  ) |> 
  ggplot(
    data = _,
    aes(x = avg_pct_chng,y = tech)
  ) + 
  facet_wrap(~gender) +
  geom_vline(
    xintercept = 0,
    color = "gray",
    linetype = "dashed"
  ) +
  geom_quasirandom() + 
  geom_point(
    data = milan_anno,
    color = "red"
  ) +
  geom_label_repel(
    data = milan_anno,
    aes(label = lbl),
    color = "red",
    size = 9 / .pt,
    direction = "y",
    force_pull = 0.5
  ) +
  labs(
    x = "Average percent change in fastest round time (relative to qualification)",
    y = NULL,
    title = "Milan Olympic Sprint Course Slowed Down Dramatically",
    subtitle = "Average percent change of fastest round time relative to best qualifying time",
    caption = "statisticalskier.netlify.app"
  ) +
  scale_x_continuous(
    breaks = seq(-0.04,0.1,by = 0.02),
    labels = scales::label_percent()
  ) + 
  scale_y_discrete(
    expand = expansion(add = c(0.5,0.5))
  )  + 
  theme(
    plot.caption = element_text(color = "grey",face = "italic")
  )
```