---
title: "Ruka 10k Classic Predictions"
author: "Joran Elias"
date: "2023-11-24"
categories: [world cup,distance,classic,predictions]
draft: false
---

```{r pkgs,message=FALSE}
library(tidyverse)
library(statskier2)
library(gt)
library(gtExtras)
library(ggpubr)

theme_set(theme_bw())
```

## Predictions
Another set of predictions, this time for the first distance race. I haven't changed the methodology at all, yet.

~~It's predicting Jessie Diggins to finish second, which seems like a moderately bold take. We'll see!~~

Ha ha! I accidentally used the start list from the women's sprint instead of the 10k classic, no wonder 
the predictions looked a little wonky! I replaced them with ones generated from the correct start list.

```{r pred-data}
mens_ruka_preds <- 
  readr::read_csv(file = '2023-11-25_men_cl_dst_ruka.csv',
                  show_col_types = FALSE)
womens_ruka_preds <- 
  readr::read_csv(file = '2023-11-25_women_cl_dst_ruka.csv',
                  show_col_types = FALSE)
```

### Men's 10k Classic Predictions
```{r tbl-men}
mens_ruka_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = "Ruka 10k Classic",subtitle = "Men") |> 
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank) 
```
### Women's 10k Classic Predictions
```{r tbl-wom}
womens_ruka_preds |> 
  select(-compid) |> 
  gt() |> 
  tab_header(title = "Ruka 10k Classic",subtitle = "Women") |>
  opt_align_table_header(align = 'left') |> 
  tab_caption(caption = "") |> 
  tab_options(
    table.font.size = 9,
    ihtml.active = TRUE,
    ihtml.use_pagination = TRUE,
    ihtml.use_compact_mode = TRUE,
    ihtml.use_highlight = TRUE
  ) |> 
  cols_label(
    fisid = 'FIS Code',
    name = 'Name',
    page_rank = 'Predicted PR',
    page_rank_pos = 'Predicted Finish') |> 
  fmt_number(columns = page_rank,decimals = 4) |> 
  gt_plt_bar(column = page_rank)
```

## Update
```{r db-con}
statskier_connect()
v_dst_maj_int <- dplyr::tbl(..statskier_pg_con..,'v_distance_maj_int')
```

```{r race-data}
ruka_dst_res <- v_dst_maj_int |> 
  filter(eventid %in% c(19176,19177)) |> 
  collect()

fis_pts <- readRDS(file = '~/Documents/code/official-fis-points-lists/output/fis_pts_list.rds') |> 
  select(listid = Listid,
         listname = Listname,
         listnum = ListNum,
         compid = Competitorid,
         year = year,
         calc_date = Calculationdate,
         dst_points = DIpoints) |> 
  filter(year >= 2020)

men_comb <- ruka_dst_res |> 
  filter(gender == 'Men') |> 
  select(date,compid,rank) |> 
  mutate(date = as.Date(date)) |> 
  full_join(mens_ruka_preds,by = 'compid') |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(
    fis_pts_rank = rank(dst_points,ties.method = 'min')) |> 
  arrange(rank)

wom_comb <- ruka_dst_res |> 
  filter(gender == 'Women') |> 
  select(date,compid,rank) |> 
  mutate(date = as.Date(date)) |>
  full_join(womens_ruka_preds,by = 'compid') |> 
  left_join(
    fis_pts,
    join_by(compid == compid,closest(date > calc_date))) |> 
  mutate(
    fis_pts_rank = rank(dst_points,ties.method = 'min')) |> 
  arrange(rank)
```

Alright, let's see how we did, starting with the men:

```{r men-perf}
men_comb |> 
  select(compid,rank,page_rank_pos,fis_pts_rank) |> 
  pivot_longer(cols = c(page_rank_pos,fis_pts_rank),names_to = 'grp',values_to = 'val') |> 
  mutate(grp = if_else(grp == 'fis_pts_rank','FIS Sprint Points List','PR Method')) |> 
  ggplot(data = _,aes(x = rank,y = val)) + 
  facet_wrap(~grp,ncol = 1) + 
  geom_point() + 
  geom_abline(slope = 1,intercept = 0,color = 'blue') + 
  stat_cor(aes(label = after_stat(..r.label..))) +
  ggtitle(label = "Men's Ruka 10k Classic Predicted vs True Athlete Finishing Order",
          subtitle = "Athletes ordered by most recent FIS Sprint points list (top) & my PR method (bottom)") +
  labs(x = 'True Finishing Place',y = 'Predicted Finishing Place')
```

Again, my predictions did somewhat better than my chosen baseline here. I'm beginning to notice a 
pattern where my predictions do particularly well for the very top & bottom of the fields, while
things are more similar between the two in the middle. 

How about the women:

```{r wom-perf}
wom_comb |> 
  select(compid,rank,page_rank_pos,fis_pts_rank) |> 
  pivot_longer(cols = c(page_rank_pos,fis_pts_rank),names_to = 'grp',values_to = 'val') |> 
  mutate(grp = if_else(grp == 'fis_pts_rank','FIS Sprint Points List','PR Method')) |> 
  ggplot(data = _,aes(x = rank,y = val)) + 
  facet_wrap(~grp,ncol = 1) + 
  geom_point() + 
  geom_abline(slope = 1,intercept = 0,color = 'blue') + 
  stat_cor(aes(label = after_stat(..r.label..))) +
  ggtitle(label = "Women's Ruka 10k Classic Predicted vs True Athlete Finishing Order",
          subtitle = "Athletes ordered by most recent FIS Sprint points list (top) & my PR method (bottom)") +
  labs(x = 'True Finishing Place',y = 'Predicted Finishing Place')
```

We did slightly better here, but not as much better as with the men. Unlike with the men, there isn't
a big difference with the performance at the front of the field, but my predictions did do noticeably
better with the bottom of the field.

