---
title: "Nation Diversity (Again)"
author: "Joran Elias"
date: "2025-03-13"
categories: [world cup]
draft: false
---

```{r pkgs,message=FALSE}
library(tidyverse)
library(ggokabeito)
library(statskier2)
library(slider)

statskier_connect()

theme_set(
  theme_minimal() + 
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      plot.title.position = "plot",
      strip.text = element_text(hjust = 0,size = rel(1.1)),
      legend.position = "bottom",
      legend.direction = "horizontal"
    )
)
```

```{r}
rel_div <- function(.data,cutoff){
  .data |> 
    filter(rank <= cutoff) |> 
    mutate(
      max_shannon = log(n_distinct(nation)),
      .by = c(gender,season)
    ) |> 
    summarise(
      n = n(),
      .by = c(season,gender,nation,max_shannon)
    ) |> 
    mutate(
      p = n / sum(n),
      .by = c(season,gender)
    ) |> 
    summarise(
      s = -sum(p*log(p)),
      .by = c(season,gender,max_shannon)
    ) |> 
    mutate(
      s_rel = s / max_shannon,
      date = season_to_date(season)
    ) |> 
    arrange(gender,season) |> 
    mutate(
      s_roll = slider::slide_dbl(.x = s,.f = mean,.before = 4L),
      s_rel_roll = slider::slide_dbl(.x = s_rel,.f = mean,.before = 4L),
      .by = gender
    ) |> 
    mutate(
      top_n = glue::glue("Top {cutoff}")
    )
}
```

```{r}
v_dst_maj_int <- dplyr::tbl(..statskier_pg_con..,"v_distance_maj_int")
v_spr_maj_int <- dplyr::tbl(..statskier_pg_con..,"v_sprint_maj_int")

dst <- 
  v_dst_maj_int |> 
  filter(!is.na(rank) & !is.na(nation)) |> 
  collect() |> 
  mutate(
    nation = toupper(nation),
    nation = 
      case_when(
        nation %in% c("RSF","ROC","SOV","RUS") ~ "SOV/RUS",
        nation %in% c("ESP","SPA") ~ "ESP",
        nation %in% c("CZE","TCH") ~ "CZE",
        nation %in% c("GER","DDR","BRD","GDR") ~ "GER",
        TRUE ~ nation),
    type = "Distance"
  )

spr <- 
  v_spr_maj_int |> 
  filter(!is.na(rank) & !is.na(nation)) |> 
  collect() |> 
  mutate(
    nation = toupper(nation),
    nation = 
      case_when(
        nation %in% c("RSF","ROC","SOV","RUS") ~ "SOV/RUS",
        nation %in% c("ESP","SPA") ~ "ESP",
        nation %in% c("CZE","TCH") ~ "CZE",
        nation %in% c("GER","DDR","BRD","GDR") ~ "GER",
        TRUE ~ nation),
    type = "Sprint"
  )

dst_spr <- 
  bind_rows(
    dst |> select(type,eventid,season,date,primary_tag,gender,compid,nation,rank),
    spr |> select(type,eventid,season,date,primary_tag,gender,compid,nation,rank)
  )
```

```{r}
only_owg_wsc <- 
  dst_spr |> 
  mutate(
    wsc_owg_flag = if_else(primary_tag %in% c("wsc","owg"),"y","n")
  ) |> 
  summarise(
    all_wsc_owg = all(wsc_owg_flag == "y"),
    .by = nation
  ) |> 
  filter(all_wsc_owg)
```

```{r}
dst_spr <- 
  anti_join(
    dst_spr,
    only_owg_wsc,
    by = "nation"
  ) 
```

The recent Trondheim World Championships once again led to lots of discussion of
Norway's dominance of the sport, particularly on the men's side, less so for the
women. This is an evergreen data topic so I thought I'd _again_ provide some 
visuals for historical context.

The data I'm working with is WC, TdS, WSC & OWG _individual_ results stretching back to 1980. No relays! 
One of the tricky things about dealing with different nations across such long 
time periods is that political boundaries change. In what follows I've made two
adjustments. The first I consider fairly reasonable, and frankly almost necessary: 
combining the Soviet athletes from the 80's with the post-Soviet breakup Russian 
athletes into a single "nation". Obviously, the USSR is not the same nation as
current day Russia, but combining the two seems like a reasonable way to track
"Russian" skiers over multiple decades. Just keep in mind that the 80's numbers 
surely included some athletes that today would be Estonian, Kazakh or something.

The second is combining both East & West Germany from the 80's with present day
unified Germany. I don't particularly _like_ this one (I actually dislike it on many levels) 
but it doesn't have a large impact on the overall message in the plots I'm making
and it greatly simplifies some technical details of actually designing the plots.

```{r}
n_nations <- 
  bind_rows(
    dst_spr |> 
      filter(rank <= 3) |> 
      summarise(
        n_nations = n_distinct(nation),
        .by = c(gender,season)
      ) |> 
      mutate(n_rank = "Top 3"),
    dst_spr |> 
      filter(rank <= 5) |> 
      summarise(
        n_nations = n_distinct(nation),
        .by = c(gender,season)
      ) |> 
      mutate(n_rank = "Top 5"),
    dst_spr |> 
      filter(rank <= 10) |> 
      summarise(
        n_nations = n_distinct(nation),
        .by = c(gender,season)
      ) |> 
      mutate(n_rank = "Top 10")
  ) |> 
  mutate(
    date = season_to_date(season),
    n_rank = factor(n_rank,levels = c("Top 3","Top 5","Top 10"),ordered = TRUE)
  )
```

Anyway, let's start very simply with a plot of the number of different nations
represented in the top 3, top 5 & top 10 for each season:

```{r}
n_nations |> 
  ggplot(
    data = _,
    aes(x = as.Date(date),y = n_nations,color = gender)
  ) + 
  facet_wrap(~n_rank,nrow = 1) +
  geom_line(linewidth = 0.7) + 
  labs(x = NULL,y = "# nations",color = NULL) +
  ggtitle(
    label = "Nations represented in WC, TdS, WSC & OWG competitions each season"
  ) +
  scale_y_continuous(
    breaks = seq(0,30,by = 2)
  ) + 
  scale_color_okabe_ito()
```

More nations until about 2010 and then it's dropped a little since. Of course,
this is very simplistic and doesn't account for changes in the number of nations
competing. The breakup of the USSR alone "increased diversity" at the top of the
results simply by creating a bunch of new nations.

```{r}
div30 <- rel_div(dst_spr,cutoff = 10)
div10 <- rel_div(dst_spr,cutoff = 5)
div03 <- rel_div(dst_spr,cutoff = 3)
```

```{r}
div_comb <- 
  bind_rows(
    div03,
    div10,
    div30
  ) |> 
  mutate(
    top_n = factor(top_n,levels = c("Top 3","Top 5","Top 10"),ordered = TRUE)
  )
```

Instead, let's use an actual measure of diversity ([Shannon index](https://en.wikipedia.org/wiki/Diversity_index)), scaled relative to
the maximum possible diversity given the number of nations competing:

```{r}
div_comb |> 
  filter(date >= as.Date("1980-01-01")) |> 
  ggplot(
    data = _,
    aes(x = as.Date(date),y = s_rel_roll,color = gender,group = gender)
  ) + 
  facet_wrap(~top_n) +
  geom_line(linewidth = 0.7) +
  labs(x = NULL,y = "Relative diveristy",color = NULL) +
  ggtitle(
    label = "Rolling average relative Shannon index"
  ) +
  scale_y_continuous(
    labels = scales::label_percent()
  ) +
  scale_color_okabe_ito()
```

I've used a rolling average here to make things a little more easy to read, given
the wild swings we're seeing.

This looks much worse. Two things stand out: the wild swings in diversity in the
women's field and the precipitous drop in diversity overall in the men's field in
the last 10-15 years.

Note that the diversity drop for the men long predates Russia's absence from the 
World Cup.

The fact that the wild swings for the women's field persists in the diversity 
measure for the top 5 & top 10 suggests that single athlete phenomenon (e.g. Justyna
Kowalczyk) are only partially responsible. Still, the late 90's dip is probably 
partially due to Elena Vaelbe, and the late 10's dip is probably Therese Johaug.

```{r}
pod_counts <- 
  dst_spr |> 
  filter(rank <= 3) |> 
  summarise(
    n_pods = n(),
    .by = c(gender,season,nation)
  )
```

Just to round things out, let's get a little more granular and look at the specific
nations at play here:

```{r}
bar_data <- 
  pod_counts |> 
  slice_max(
    order_by = n_pods,
    n = 5,
    by = c(gender,season)
  ) |> 
  mutate(
    nation = if_else(
      nation %in% c("NOR","SWE","SOV/RUS","ITA","FIN","GER"),
      nation,
      "Other"
    )
  ) |> 
  mutate(
    tot = sum(n_pods),
    .by = nation
  ) |> 
  arrange(desc(tot)) |> 
  mutate(
    nation = fct_inorder(nation),
    nation = fct_relevel(nation,"Other",after = 6),
    season_int = as.integer(substr(season,6,9))
  )

x_vals <- sort(unique(bar_data$season_int))

bar_data |> 
  filter(season_int >= 1980) |> 
  ggplot(
    data = _,
    aes(x = season_int,y = n_pods,fill = nation)
  ) + 
  facet_wrap(~gender,ncol = 1) + 
  geom_col(position = position_fill()) + 
  geom_hline(
    yintercept = c(0.25,0.5,0.75),
    color = "white",
    linewidth = 0.1
  ) +
  geom_segment(
    data = data.frame(
      x = seq(1980,2020,by = 10),
      xend = c(seq(1989,2020,by = 10),2025),
      y = -0.01,
      yend = -0.01,
      gender = "Women"
    ),
    aes(x = x,xend = xend,y = y,yend = yend,fill = NULL)
  ) +
  scale_x_continuous(
    breaks = c(1985,1995,2005,2015,2023) - 0.5,
    labels = c("1980's","1990's","2000's","2010's","2020's"),
    expand = expansion(mult = c(0,0))
  ) +
  scale_y_continuous(
    labels = scales::label_percent()
  ) +
  scale_fill_okabe_ito() + 
  guides(
    fill = guide_legend(nrow = 1)
  ) +
  labs(x = NULL,y = NULL,fill = NULL) + 
  ggtitle(
    label = "Top 3 finishes by nation in WC, TdS, WSC & OWG",
    subtitle = "1980's GER represents both East & West Germany combined"
  ) +
  theme(
    axis.text.x = element_text(vjust = 4)
  )
```

Norway has not always _completely_ dominated the top of the sport, but I really only
see two (and a half) exceptions to Norwegian dominance in this plot: the (probably artificially 
assisted) dominance of the Russian women in the 90's, and the Swedish men in the 
late 80's. The half an exception is the current surge of the Swedish women.

Finally, for completeness, here's what the "Other" category looks like if we
break them out and plot them on their own:

```{r}
bar_data <- 
  pod_counts |> 
  slice_max(
    order_by = n_pods,
    n = 5,
    by = c(gender,season)
  ) |> 
  filter(!nation %in% c("NOR","SWE","SOV/RUS","ITA","FIN","GER")) |> 
  mutate(
    nation = if_else(
      nation %in% c("USA","CZE","POL","FRA","SUI","KAZ"),
      nation,
      "Remaining"
    )
  ) |> 
  mutate(
    tot = sum(n_pods),
    .by = nation
  ) |> 
  arrange(desc(tot)) |> 
  mutate(
    nation = fct_inorder(nation),
    nation = fct_relevel(nation,"Remaining",after = 6),
    season_int = as.integer(substr(season,6,9))
  )

bar_data |> 
  filter(season_int >= 1980) |> 
  ggplot(
    data = _,
    aes(x = season_int,y = n_pods,fill = nation)
  ) + 
  facet_wrap(~gender,ncol = 1) + 
  geom_col(position = position_fill()) + 
  geom_hline(
    yintercept = c(0.25,0.5,0.75),
    color = "white",
    linewidth = 0.1
  ) +
  geom_segment(
    data = data.frame(
      x = seq(1980,2020,by = 10),
      xend = c(seq(1989,2020,by = 10),2025),
      y = -0.01,
      yend = -0.01,
      gender = "Women"
    ),
    aes(x = x,xend = xend,y = y,yend = yend,fill = NULL)
  ) +
  scale_x_continuous(
    breaks = c(1985,1995,2005,2015,2023) - 0.5,
    labels = c("1980's","1990's","2000's","2010's","2020's"),
    expand = expansion(mult = c(0,0))
  ) +
  scale_y_continuous(
    labels = scales::label_percent()
  ) +
  scale_fill_okabe_ito() + 
  guides(
    fill = guide_legend(nrow = 1)
  ) +
  labs(x = NULL,y = NULL,fill = NULL) + 
  ggtitle(
    label = "Breakdown of 'Other' category",
    subtitle = "Remaining nations mostly comprised of EST, SLO, CAN & AUT"
  ) +
  theme(
    axis.text.x = element_text(vjust = 4)
  )
```

Here we can start to identify the impacts of specific individual skiers like 
Smirnov, Kowalczyk and Diggins.